---
title: "RMS Titanic"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
summary(df_titanic)
```

**Observations**:

-   Class, Sex, Adult, Survived, n

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**

-   
    -   The number of people on board according to the Wiki is 2224. While the number reported by the data set is 2201.
-   Are there any differences?
    -   Yes, a difference by 23
-   If yes, what might account for those differences?
    -   The difference might be due to the uncounted stowaways and maybe unborn children among the women.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`

df_titanic %>%
  filter(Survived == "Yes") %>%
  group_by(Class, Sex) %>%
  ggplot(aes(fill=Sex, y=n, x=Class)) + 
    geom_bar(position="stack", stat="identity") + 
    labs(title = "Survival Count by Class and Sex", x = "Class", y = "# of People Survived")

```

**Observations**:

-   In both the 1st and 2nd class, the survival rate of females are greater than the survival rate of males.
    -   This is likely because the crew prioritized saving females and children when there are limited life boats
-   The crew has mostly male survivors
    -   This may be because there were very few female staff.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  group_by(Class, Sex) %>%
  ggplot(aes(fill=Sex, y=Prop, x=Class)) + 
    geom_bar(position="dodge", stat="identity") + 
    labs(title = "Survival Proportion by Class and Sex - FISHY", x = "Class", y = "Proportion")
```

**Observations**:

-   It seems like 1st class had a higher survival rate proportionally as compared to the other classes.
-   Is there anything *fishy* going on in your plot?
    -   The bars in this graph for proportions adds up to over 1, which seems incorrect because proportions should range from 0-1. This would mean that more people than the number of people on the ship survived.
        -   NOTE: this was addressed by changing *position="stack"* to *position="dodge".*
    -   It looks like everyone survived! That can't be right?

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  group_by(Class, Sex) %>%
  ggplot(aes(fill=Sex, y=Prop, x=Class)) + 
    geom_bar(position="dodge", stat="identity") + 
    labs(title = "Survival Proportion by Class and Sex", x = "Class", y =   "Proportion") + 
    facet_grid("Age")
```

**Observations**:

-   All the children in 1st and 2nd class survived. The same could not be said for the 3rd class kids.
-   For the adults, The survival rate of females are much higher than that of the males.
-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   The graph is q4 was fishy because it used the large percentage from either adults/child. This is shown by the FISHY graph matching up with the 1st, 2nd, and 3rd class of the child's graph, and the adult part matching up with the Crew part.
    -   By separating out the categories, we're able to avoid this issue.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
