---
title: "Gapminder"
author: "Jacqueline Zeng"
date: 2020-2-26
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|--------------|----------------------------|-------------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
# glimpse(gapminder)
head(gapminder)
# summary(gapminder)
```

**Observations**:

-   country, continent, year, lifeExp, pop, gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- max(pull(gapminder, year))
year_min <- min(pull(gapminder, year))

print(year_max)
print(year_min)
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent

# Attempt 1: I tried making a boxplot but the outliers are too big and prevent the graph from being readable
df_q2 <- gapminder %>% 
  filter(year == year_min) %>% 
  group_by(continent)

df_q2 %>%
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  labs(title = "Continent vs GDP per Capital for 1952", x = "Continent", y = "GDP per Capita")

# Attempt 2: Remove the outliers and replot the boxplot

# Find outliers
out <- boxplot.stats(df_q2$gdpPercap)$out
out_ind <- which(df_q2$gdpPercap %in% c(out))
print(df_q2[out_ind, ])

# Create dataset with outliers removed
df_q2_without_outliers <- df_q2[-c(out_ind), ] 

# Boxplot without outliers
df_q2_without_outliers %>% 
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  labs(title = "Continent vs GDP per Capita for 1952 (outliers removed)", x = "Continent", y = "GDP per Capita")
```

**Observations**:

-   Europe has the widest spread of GDP per capita of the 4 continents for the year 1952
-   Oceania's continent category was removed after filtering out outliers
-   Africa and Asia have similar GDP medians, but Asia's upper quartiles are higher. America's median GDP is higher than Africa and Asia but less than Europe's

**Difficulties & Approaches**:

-   I first filtered the dataset to match the min_year value and grouped the countries by continents. I then plotted a boxplot based on this data. The resulting boxplot (first output in the above code section) was difficult to understand as the outliers were extremely high and squashed the boxplots into a thin sliver at the bottom of the plot. This was not very readable.
-   I addressed this by identifying the outliers, removing them from the dataset, and replotting a boxplot.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
out <- boxplot.stats(df_q2$gdpPercap)$out
out_ind <- which(df_q2$gdpPercap %in% c(out))
print(df_q2[out_ind, ])
```

**Observations**:

-   Identify the outlier countries from q2
    -   Australia, Bahrain, Canada, Denmark, Kuwait, Netherlands, New Zealand, Norway, Switzerland, United Kingdom

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
df_q4 <- gapminder %>% 
  filter(year == c(year_min,year_max)) %>% 
  group_by(continent)

# Get outliers
out <- boxplot.stats(df_q4$gdpPercap)$out
out_ind <- which(df_q4$gdpPercap %in% c(out))
print(df_q4[out_ind, ])

# Plot both graphs on same plot
df_q4 %>%
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  labs(title = "Continent vs GDP per Capita for 1952", x = "Continent", y = "GDP per Capita") + facet_grid(cols = vars(year)) +
  geom_point(
    data = . %>% filter(country %in% c("Australia", "Bahrain", "Canada", "Denmark", "Kuwait", "Netherlands", "New Zealand", "Norway", "Switzerland", "United Kingdom")),
    mapping = aes(color = country),
    size = 2
  )
df_q4

```

**Observations**:

-   Africa's median GDP is around the same for 1952 and 2007, but the it's outliers are higher
-   Asia, Europe, America, and Oceania all experienced some degree of median GDP rises and widening of the range of GDPs.
-   The upper outliers for 2007 are on average more extreme

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
# GOAL: What is the life expectancy trend from the first year recorded to the latest year recorded?

# Get outliers
out <- boxplot.stats(df_q4$lifeExp)$out
out_ind <- which(df_q4$lifeExp %in% c(out))
print(df_q4[out_ind, ])

# Plot both graphs on same plot
df_q4 %>%
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  labs(title = "Life Expectancy by Continent", x = "Continent", y = "Life Expectancy (Years)") + facet_grid(cols = vars(year))

```

-   The median life expectancy for all continents rose throughout the years
-   Asia's rise in life expectancy is the biggest - the median rose from around 45 years to 73 years old
-   Africa's whiskers grew longer while all the other continent's whiskers shrunk

```{r q5-task2}
## TASK: Your second graph
# GOAL: Trend with the population
# Get outliers
out <- boxplot.stats(df_q4$pop)$out
out_ind <- which(df_q4$pop %in% c(out))
print(df_q4[out_ind, ])

# Plot both graphs on same plot
df_q4 %>%
  ggplot(aes(x = continent, y = pop)) +
  geom_boxplot() +
  labs(title = "Population by Continent", x = "Continent", y = "Population") + facet_grid(cols = vars(year))

```

-   Asia has the biggest outliers in terms of population
-   Asian has the biggest spread of population
-   Europe and Oceania didn't have significant changes
-   Africa and America's spread and outliers did not change was significantly as Asia, but the outliers got bigger, and the spread also increased.

```{r q5-task3}
## TASK: Your third graph
# GOAL: deep dive into Asia's population

df_asia <- df_q4 %>% 
  filter(continent == "Asia")

# Get outliers
out <- boxplot.stats(df_asia$pop)$out
out_ind <- which(df_asia$pop %in% c(out))
print(df_asia[out_ind, ])

df_asia %>%
  ggplot(aes(x = continent, y = pop)) +
  geom_boxplot() +
  labs(title = "Asia Population", x = "Year", y = "Population") + facet_grid(cols = vars(year)) +
  geom_point(
    data = . %>% filter(country %in% c("Bangladesh", "China", "India", "Indonesia", "Japan", "Pakistan")),
    mapping = aes(color = country),
    size = 2
  )

```

-   In a deeper dive into Asia's population trend from the first year to the last year of the dataset
-   China is the most populous, with India behind it. These are the two biggest outliers
